Reasonably Programmable Literal Notation

### 2018-03-07 icfp18/syntax-icfp18.tex
more tweaks to introgit push!

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}. Briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches, like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at other types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing. Only existing composite forms can therefore be repurposed.
\item An approach that came close was Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}, but it fell short with regard to the six reasoning principles just discussed in ways that we will return to and in any case, the mechanism was designed for a simple monomorphic, nominally-typed language and relied critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple types and arrow types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Brief Overview} To finish up our discussion of the K query example above, let us give a brief overview of the TLM-based solution outlined in Figure \ref{fig:K-tsm-example}. On lines 1-5, we define a TLM named \li{$kq}. On Line 9, we apply this TLM. We can reason abstractly about this program as follows:
\begin{enumerate}[leftmargin=15pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms, and on constructing a TLM lexer and parser in the next section. For now, it suffices to say that our design goal is to provide a mechanism where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: The external dependencies of the generated expansion are explicitly specified. Here, Lines 4-5 specify that expansions generated by \li{$kq} require the module \li{KQuery}. The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site. By making dependencies explicit, we can be sure that the expansion does not rely on the fact that, e.g., \li{w} is in scope at the application site. 
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to editor services downstream of the expander, and ultimately to the programmer using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are green).
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by requesting just the type annotations on the spliced segments.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes \emph{regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regular expressions are  common in fields like bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify the constructors \li{Regex.Or} and \li{Regex.Str}. Even allowing for this shorthand, however, this notation for constructing regexes is rather unwieldy. Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.

>>>>

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}. Briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches, like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at other types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing. Only existing composite forms can therefore be repurposed.
\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec. \ref{sec:existing-approaches}. In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}. On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation. On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}. We can reason abstractly about this program as follows:
\begin{enumerate}[leftmargin=12pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section. For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others. The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site. This completely relieves clients from needing to consider expansion-internal dependencies when naming variables. %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to language services downstream of the expander, e.g. editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are green).
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by reference only to the type annotations on the spliced segments, not the full expansion.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with this shorthand, however, this notation for constructing regexes is rather unwieldy. Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.
